### 2023-03-14 paper.tex
describe dmr stuff

\begin{figure*}[t]
    \begin{center}
        \includegraphics[width=0.8\textwidth]{figures/Fig_benchmark.png}
    \end{center}
    \caption{\small \textbf{Benchmark of our methods on single-cell multi-omic data of cells of the murine forebrain}.\\
    \textbf{(A)} Cell labels based on clustering of single-cell transcriptomes from \citet{kremer_scnmt}.
    \textbf{(B, C)} Exemplary UMAPs obtained when analyzing the data set with our proposed methods (B) or conventional methods based on genome tiling (C).
    \textbf{(D, E)} Mean neighbor score \textbf{(D)} obtained after analysing single-cell methylomes with different combinations of methods.
    Either 100~kb genomic tiles or VMRs were subjected to PCA and UMAP.
    Tiles or VMRs were pre-processed as proposed in this work (shrunken mean of residuals subjected to iterative PCA, labeled "pre-processing: ours") or using conventional methods (PCA on lightly imputed methylation percentages, based on \citep{luo2017single}).
    The full 1566-cell data set was sub-sampled to simulate smaller data sets (x axis).
    A higher neighbor score implies better separation of cell types inferred from single-cell transcriptomes of the same cells \textbf{(A)}.
    }
    \label{fig:score}
\end{figure*}

@@

\phantomsection
\subsection{Detecting DMRs between oligodendrocytes and neuroblasts}

\begin{figure*}[t]
    \begin{center}
        \includegraphics[width=.7\textwidth]{figures/Fig_DMRs.png}
    \end{center}
    \caption{\small \textbf{Detection of differentially methylated regions between neural stem cells and oligodendrocytes.}\\
    \textbf{(A)} DMRs detected between neural stem cells (NSCs) and oligodendrocytes (ODs) from \citep{kremer_scnmt} (top) and DMRs detected in the same data with randomly permuted cell labels (bottom, used to estimate FDR and determine adjusted p-values).
    \textbf{(B)} Enrichment of GO terms associated with DMRs lowly methylated in ODs (top) or NSCs (bottom).
    Depicted are the top 5 GO terms of the "Biological Process" (GO:BP) and "Cellular Component" (GO:CC) GO category, and their binomial test p-value and enrichment, as reported by GREAT \citep{mclean2010great}.
    \textbf{(C)} Mean methylation of NSCs and ODs at an exemplary DMR.
    Each point corresponds to a cell, black lines denote the median.
    \textbf{(D)} Detailed view of the DMR (yellow dashed rectangle) from \textbf{C} in pseudo-bulk samples consisting of NSCs or ODs.
    Vertical bars represent CpG sites.
    }
    \label{fig:dmr}
\end{figure*}

Next, we used the single-cell multi-omics data set \citep{kremer_scnmt} to assess the ability of our DMR detection approach.
We first selected two cell populations from the murine ventricular-subventricular zone, neuroblasts and oligodendrocytes, and then used the method depicted in Fig.~\ref{fig:dmrscan} to identify regions with differing methylation.
Fig.~\ref{fig:dmr} gives an overview of all identified DMRs and the relation of the adjusted p-value to the number of cells included in a DMR, and the difference in methylation between the cell types.
Logically, if a DMR contains few cells, the methylation difference between cell types must be higher to obtain a low (and thus more significant) p-value.
A low p-value is maintained, although the methylation difference decreases if the number of cells in the DMR increases.

An oligodendrocyte DMR stretching across Mbp was chosen to assess the methylation patterns of an individual DMR (Fig.~\ref{fig:dmr}B-C).
Mbp is highly expressed in oligodendrocytes as it is essential for the myelination of neuronal axons.
The DMR is highly methylated in neuroblasts and lowly methylated in oligodendrocytes (Fig.~\ref{fig:dmr}B-C), including more substantial demethylation in oligodendrocytes at the position of Mbp (Fig.~\ref{fig:dmr}C).
Low methylation at this locus supports the expression of Mbp.

Nevertheless, this is not the only indicator of the biological relevance of discovered DMRs.
GO-term enrichment shows that neuroblast DMRs occur near genes involved in the regulation of neurogenesis, neuron differentiation, and the positive regulation of neural precursor cell proliferation, all of which are essential for neuroblasts (Fig.~\ref{fig:dmr}D).
On the other side, oligodendrocyte-defining DMRs are located near genes with implications for myelin sheath, myelination, and axon ensheathment and thus are linked to oligodendrocytes.
Therefore, DMRs occur in the proximity of genes associated with oligodendrocyte or neuroblast function, respectively and are implied to regulate gene expression and contribute to fundamental cell type-specific function, which is supported by the repressive effect of CG methylation on gene expression.

@@

We have implemented the approach just described in a Python package with a command line interface, called \texttt{scbs}, which also offers a number of other functionalities for the analysis of scBS data (Figure \ref{fig:workflow}).
The starting point of such an analysis are methylation files generated by tools such as Bismark \citep{bismark} or methylpy \citep{methylpy}.
For each cell, these tools produce a text file that lists the methylation status of all CpG sites.
Since it is inconvenient to work with hundreds or thousands of text files, \texttt{scbs} provides the \texttt{prepare} command which parses these methylation files and stores their content in a compressed format that enables efficient access to all CpG sites of the genome.
\texttt{scbs prepare} is flexible and accepts all tabular input formats including Bismark .cov-files, methylpy .allc-files, or custom user-defined formats.

@@

After quality control, the user has access to the genome-wide VMR detection approach described earlier, using the \texttt{scbs scan} command.
This produces a BED file that lists the genome coordinates of VMRs.
Similarly, the user can identify DMRs after quality control with \texttt{scbs diff} using a text file listing the cells’ group assignment as input.
The command produces a BED file listing the coordinates of the determined DMRs together with the t- statistic, the corresponding cell group, and the adjusted p-value.
To finally obtain a methylation matrix analogous to a scRNA-seq count matrix, this BED file can be used as input for \texttt{scbs matrix}, which quantifies methylation at genomic intervals in all cells.
The command produces both the percentage of methylated CpG sites, as well as our proposed methylation measure (shrunken mean of the residuals) that is more robust to variation in sequencing coverage or stochastic differences in read position between cells.
We note that both \texttt{matrix} and \texttt{profile} accept any valid BED file as input, which means that the user can quantify and visualize methylation at any genomic feature of interest, such as promoters, enhancers or transcription factor binding sites, in single cells.
The obtained methylation matrix can then serve as input for common single-cell analysis methods such as dimensionality reduction and cell clustering.

>>>>

\begin{figure*}[t]
    \begin{center}
        \includegraphics[width=0.8\textwidth]{figures/Fig_benchmark.png}
    \end{center}
    \caption{\small \textbf{Benchmark of our methods on single-cell multi-omic data of cells of the murine forebrain}.\\
    \textbf{(A)} Cell labels based on clustering of single-cell transcriptomes from \citet{kremer_scnmt}.
    \textbf{(B, C)} Exemplary UMAPs obtained when analyzing the data set with our proposed methods (B) or conventional methods based on genome tiling (C).
    \textbf{(D, E)} Mean neighbor score \textbf{(D)} obtained after analysing single-cell methylomes with different combinations of methods.
    Either 100~kb genomic tiles or VMRs were subjected to PCA and UMAP.
    Tiles or VMRs were pre-processed as proposed in this work (shrunken mean of residuals subjected to iterative PCA) or using conventional methods (PCA on lightly imputed methylation percentages, based on \citep{luo2017single}).
    The full 1566-cell data set was sub-sampled to simulate smaller data sets (x axis).
    A higher neighbor score implies better separation of cell types inferred from single-cell transcriptomes of the same cells \textbf{(A)}.
    }
    \label{fig:score}
\end{figure*}

@@

\phantomsection
\subsection{Detecting DMRs between oligodendrocytes and neural stem cells}

\begin{figure*}[t]
    \begin{center}
        \includegraphics[width=.7\textwidth]{figures/Fig_DMRs.png}
    \end{center}
    \caption{\small \textbf{Detection of differentially methylated regions between oligodendrocytes and neural stem cells.}\\
    \textbf{(A)} DMRs detected between 58 oligodendrocytes (ODs) and 130 neural stem cells (NSCs) from \citep{kremer_scnmt} (top) and DMRs detected in the same data with randomly permuted cell labels (bottom, used to estimate FDR and determine adjusted p-values).
    \textbf{(B)} Enrichment of GO terms associated with DMRs lowly methylated in ODs (top) or NSCs (bottom).
    Depicted are the top 5 GO terms of the "Biological Process" (GO:BP) and "Cellular Component" (GO:CC) GO category, and their binomial test p-value and enrichment, as reported by GREAT \citep{mclean2010great}.
    \textbf{(C)} Mean methylation of NSCs and ODs at an exemplary DMR.
    Each point corresponds to a cell, black lines denote the median.
    \textbf{(D)} Detailed view of the DMR (yellow dashed rectangle) from \textbf{C} in pseudo-bulk samples consisting of NSCs or ODs.
    Vertical bars represent CpG sites.
    }
    \label{fig:dmr}
\end{figure*}

Next, we used the single-cell multi-omics data set \citep{kremer_scnmt} to evaluate our DMR detection approach.
We first selected two cell populations from the healthy murine ventricular-subventricular zone: neural stem cells (NSCs, 130 cells) and oligodendrocytes (58 cells).
Then we used the method depicted in Fig.~\ref{fig:dmrscan} to identify DMRs between these two cell types (Fig.~\ref{fig:dmr}).
To estimate the false discovery rate and obtain adjusted p-values, we repeated the DMR detection approach on a permutation of the same data set, generated by shuffling cell labels.
DMRs detected in this manner showed more modest methylation differences, and as a result many DMRs detected in the original data were assigned low adjusted p-values (Fig.~\ref{fig:dmr}A).
Gene ontology (GO) enrichment with GREAT \citep{mclean2010great} revealed that DMRs lowly methylated in oligodendrocytes are located near genes involved in myelination, the main function of oligodendrocytes.
Similarly, DMRs specifically demethylated in NSCs occur near genes involved in stem cell population maintenance.
This demonstrates that our DMR detection approach is able to identify biologically meaningful DMRs, even in scBS data sets of modest cell number.

Fig.~\ref{fig:dmr}C-D depicts an exemplary DMR, located at the gene body of myelin-basic protein (Mbp), the major component of myelin that is essential for myelination of neuronal axons \citep{mbp}.
Our results suggest that oligodendrocyte-specific gene expression of Mbp is supported by low methylation at the detected DMR.

@@

We have implemented the approach just described in a Python package with a command line interface, called \texttt{scbs}, which also offers a number of other functionalities for the analysis of scBS data (Figure \ref{fig:workflow}).
The starting point of such an analysis are methylation files generated by tools such as Bismark \citep{bismark}, methylpy \citep{methylpy} or biscuit \citep{biscuit}.
For each cell, these tools produce a text file that lists the methylation status of all CpG sites.
Since it is inconvenient to work with hundreds or thousands of text files, \texttt{scbs} provides the \texttt{prepare} command which parses these methylation files and stores their content in a compressed format that enables efficient access to all CpG sites of the genome.
\texttt{scbs prepare} is flexible and accepts all tabular input formats including Bismark .cov-files, methylpy .allc-files, or custom user-defined formats.

@@

After quality control, the user has access to the genome-wide VMR detection approach described earlier, using the \texttt{scbs scan} command.
This produces a BED file that lists the genome coordinates of VMRs.
To finally obtain a methylation matrix analogous to a scRNA-seq count matrix, this BED file can be used as input for \texttt{scbs matrix}, which quantifies methylation at genomic intervals in all cells.
The command produces both the percentage of methylated CpG sites, as well as our proposed methylation measure (shrunken mean of the residuals) that is more robust to variation in sequencing coverage or stochastic differences in read position between cells.
We note that both \texttt{matrix} and \texttt{profile} accept any valid BED file as input, which means that the user can quantify and visualize methylation at any genomic feature of interest, such as promoters, enhancers or transcription factor binding sites, in single cells.
The obtained methylation matrix can then serve as input for common single-cell analysis methods such as dimensionality reduction and cell clustering.

Finally, after annotation and exploration of the data set, the user may specify two groups of cells for DMR detection with \texttt{scbs diff}.
This command produces a BED file listing the genome coordinates, adjusted p-values, as well as several other metrics associated with DMRs.
These DMRs may then be associated with nearby genes, or used for GO enrichment with tools such as GREAT \citep{mclean2010great} to enable functional interpretation.
